\name: A flexible and data-driven $K$-process model for nucleosynthesis

### 2023-06-03 Paper/main.tex
Lots and lots of content

% Abundance and stellar macros
\newcommand{\mgfe}[0]{[{\rm Mg/Fe}]} 
\newcommand{\Acc}{A_{\rm cc}}
\newcommand{\AIa}{A_{\rm Ia}} 
\newcommand{\RIa}{R_{\rm Ia}^X}
\newcommand{\aIa}{\alpha_{\rm Ia}}
\newcommand{\acc}{\alpha_{\rm cc}}
\newcommand{\afe}[0]{[\alpha/{\rm Fe}]}
\newcommand{\femg}{[{\rm Fe}/{\rm Mg}]} 
\newcommand{\xmg}{[{\rm X}/{\rm Mg}]} 
\newcommand{\mgh}{[{\rm Mg}/{\rm H}]}
\newcommand{\feh}[0]{[{\rm Fe/H}]} 
\newcommand{\pIa}{p_{\rm Ia}^{\rm X}}
\newcommand{\pcc}{p_{\rm cc}^{\rm X}}
\newcommand{\fcc}{f_{\rm cc}}
\newcommand{\xfe}{[{\rm X}/{\rm Fe}]} 
\newcommand{\pIasun}{p_{\rm Ia, \odot}^{\rm X}}
\newcommand{\pccsun}{p_{\rm cc, \odot}^{\rm X}}
\newcommand{\logg}{\log(g)}
\newcommand{\teff}{T_{\rm eff}}
\newcommand{\kpc}{\rm \; kpc}
\newcommand{\kel}{\rm \; K}
\newcommand{\msun}{M_{\odot}}
\newcommand{\rgc}{R_{\rm GC}}
\newcommand{\msunvice}{M_{\odot}/M_{\odot \rm formed}}
\newcommand{\MZAMS}{M_{\rm ZAMS}}
\newcommand{\Msun}{M_{\odot}}
\newcommand{\qIax}{q_{\rm Ia}^{\rm X}}
\newcommand{\qccx}{q_{\rm cc}^{\rm X}}
\newcommand{\qagbx}{q_{\rm AGB}^{\rm X}}
\newcommand{\qIa}{q_{\rm Ia}}
\newcommand{\qcc}{q_{\rm cc}}
\newcommand{\qagb}{q_{\rm AGB}}
\newcommand{\Aagb}{A_{\rm AGB}}
\newcommand{\eumg}{[{\rm Eu}/{\rm Mg}]} 
\newcommand{\mgmn}{[{\rm Mg}/{\rm Mn}]} 
\newcommand{\alfe}{[{\rm Al}/{\rm Fe}]} 
\newcommand{\omg}{[{\rm O}/{\rm Mg}]} 
\newcommand{\xo}{[{\rm X}/{\rm O}]} 
\newcommand{\xh}{[{\rm X}/{\rm H}]} 
\newcommand{\angstrom}{\textup{\AA}}
\newcommand{\ejg}[1]{\textcolor{red}{EJG: #1}}

@@

\newcommand{\name}{\textsl{Foo}} % Hogg's just making this up; change it!!
\newcommand{\documentname}{\textsl{Article}}

\shorttitle{data-driven few-process model for nucleosynthesis}
\shortauthors{griffith and hogg}
\addtolength{\topmargin}{-0.4in} % trust in Hogg
\addtolength{\textheight}{0.7in} % trust in Hogg
\setlength{\parindent}{1.6em}
\renewcommand{\paragraph}[1]{\bigskip\par\noindent{\textbf{#1}}~---}
\sloppy\sloppypar\raggedbottom\frenchspacing % Trust Hogg

@@

\title{\name: A data-driven few-process model for nucleosynthesis}

@@

% \author[0000-0001-7775-7261]{David H. Weinberg}
% \affiliation{The Department of Astronomy and Center of Cosmology and AstroParticle Physics, The Ohio State University, Columbus, OH 43210, USA}
% \affiliation{The Institute for Advanced Study, Princeton, NJ, 08540, USA}

\begin{abstract}\noindent % trust Hogg
Testing

Stellar surface abundances look like they are produced by two dominant processes, one prompt and one delayed.
We analyze YYY abundance ratios (ratioed to Mg) for XXXX stars from APOGEE DR17 with a flexible two-process model, in which all element process amplitudes are free parameters.
The two-process model describes each element abundance of each star as the vector sum of a CCSN and SNIa process.
Prior work derives process amplitudes from median observed high-Ia and low-Ia abundance trends; we relax these assumptions and simultaneously fit all model parameters to the data, with only minimal constraints to keep the processes interpretable.
We compare fit parameters from this new method to those derived from prior work [that the two methods produce similar results but with XXX differences and identify similar outlier stars??].
Our flexible method---dubbed \name---tests the validity of the standard two-process assumptions, such as Mg being a pure CCSN element, and that Solar abundances are generated by a roughly half-half combination of CCSN and SNIa processes.
We find [that the assumptions of prior work are generally confirmed??].
We capitalize on the flexibility to explore the addition of a third process, agnostic about its origin.
The third process most improves [X elements? most significant for X elements?].
Most importantly, \name{} can be extended to populations at very low metallicity, and populations without two clear abundance sequences, such as the LMC/SMC, Sagittarius, and Gaia-Enceladus (??), making it a general diagnostic for enrichment; in principle, it can also be used to constrain fundamental nucleosynthetic parameters.
\end{abstract}

@@

\section*{}\clearpage
HOGG SAY: Here are some things to do or think about:
\begin{itemize}
  \item Should we think of this as a 2-process model or as a $K$-process model, where $K$ is often 2? I'm leaning towards $K$.
  \item \ejg{I like the K-process model. (This is named $N$-Process model in W22)}
  \item What are we going to do in terms of interpreting the results? These process vectors aren't exactly yields; they are something way more complicated, right?
  \item \ejg{Interpret in terms of $\fcc$. Can also make inferences about the metalicity dependence of SNIa/CCSN enrichment to given elements based on how $\qcc$ and $\qIa$ evolve with time.}
\end{itemize}

@@

After hydrogen, helium, lithium, and beryllium, essentially all other naturally occurring elements are made in stars, and the collisions of stars.
We are literally made of star stuff.
Stellar surface abundances---the abundances measured by taking a spectrum of a stellar photosphere---are thought to deliver a relatively unprocessed (for most elements) record of the element abundances in the gas from which the star formed (though see \ejg{Add citations}).
These birth abundances were set by a combination of nucleosynthetic processes involved in making heavy atomic nuclei, and astrophysical processes involved in delivering atoms from stellar interiors to star-formation sites \citep[e.g.,][]{johnsonja2020}.
Thus nuclear physics and a wide swath of astrophysics are critically intertwined in our understanding of stellar surface abundances.
That's important, and motivates a lot of theoretical, experimental, and observational work in nuclear physics and astrophysics.

At the present day, stellar surface abundances are not very well explained by \textsl{ab initio}, physics-driven models.
Yields vary from set to set, as they are dependent on progenitor properties and explosion assumptions \citep[e.g.,][]{rybizki2017, griffith2021b}. 
The wide parameter space of progenitor and supernovae models coupled with uncertainties in reaction rates and explosion physics hinder the creation of an accurate nucleosynthetic model from theory alone.
In the long run, it is incumbent upon us to understand these issues and correct the assumptions or calculations underlying our nucleosynthetic and astrophysical models.
In the short run, however, we gather data---tens of millions of abundance measurements on millions of stars in different astronomical surveys such as RAVE, SEGUE, LAMOST, Gaia-ESO, APOGEE, GALAH, and H3 \citep{steinmetz2006, yanny2009, gilmore2012, desilva2015, luo2015, majewski2017, conroy2019}.
This raises the question: \emph{Can we take a data-driven approach to nucleosynthesis?}

In this \documentname{}, we build a purely data-driven model for the surface element abundances observed in stars.
We treat each star as being a linear combination of two nucleosynthetic processes, one of which is primarily responsible for the $\alpha$-element Mg \citep[core collapse supernovae (CCSN), e.g.][]{andrews2017}, and one of which is primarily \emph{not} (Type-Ia supernovae (SNIa)).
Beyond that up-front assumption, we try to be agnostic about how the elements are produced.

We build upon the work of \citet[][hereafter G22]{griffith2019, griffith2022} and \citet[][hereafter W22]{weinberg2019, weinberg2022}, who used the bimodality in [Mg/Fe] vs. [Fe/H] \citep[e.g.,][]{fuhrmann1998, bensby2003, adibekyan2012} to separate stars into populations with high and low SNIa enrichment. Using the median [X/Mg] vs. [Mg/H] abundance trends, these works explain data from the \textsl{GALAH}\footnote{GALAH = GALactic Archaeology with HERMES.} and SDSS-IV \textsl{APOGEE}\footnote{APOGEE = Apache Point Observatory Galactic Evolution Experiment, part of the Sloan Digital Sky Survey} surveys, respectively, with a two-process model. Because the median abundance trends in [X/Mg] vs. [Mg/H] space are largely insensitive to aspects of chemical evolution, such as outflows and variations in star formation history \citep{weinberg2019}, the population abundance trends are set by the nucleosynthetic processes and can be used to empirically constrain Galactic enrichment. To fit the two-process model to survey data, G22 and W22 assume that Mg is purely produced by CCSN, Fe is produced in equal amounts by CCSN and SNIa, and that the CCSN/SNIa yields of Mg and Fe are metallicity independent. While the first assumption is firmly grounded in nucleosynthetic theory \citep[e.g.,][]{andrews2017, rybizki2017}, the others are based on APOGEE abundance patterns and may not be realistic constraints.

Beyond the two-process model, many elements have contributions from additional nucleosynthetic processes, such as the rapid ($r$) and slow ($s$) neutron capture processes \citep{arlandini1999, bisterzo2014} in asymptotic giant branch (AGB) stars \citep[e.g.,][]{simmerer2004, karakas2016} and merging neutron stars \citep{kilpatrick2017}, or atypical supernovae explosion. After predicting stellar abundances from $\feh$ and $\mgfe$, \citet{ting2022} identify correlated abundance residuals that are unexplained by observational uncertainties, indicative of additional nucleosynthetic processes that standard disk CCSN and SNIa enrichment cannot explain. Results from G22 and W22 support this conclusion, and both works attempt to add additional processes to their models to account for non-CCSN and non-SNIa enrichment, though in a very restrictive manner.

\ejg{Add paragraph about how little work has been done to optimize stellar abundances to solve for enrichment processes in a more data-driven way. \citet{casey2019} apply a clustering algorithm to GALAH DR2 data, identifying latent factors corresponding to 6 processes. Difficult to interpret and breaks down with more data.}

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data. The benefit we will gain from this is better performance at fitting the data.
The cost we might pay is a possible reduction in interpretability.
We'll come back to all that at the end.

\ejg{Placeholder. We attempt to find the intersection between reliable facts about nucleosynthesis and good abundance measurements. We are trying to build an edifice of nucleosynthesis on a small number of facts. Because Mg is well-measured and well-understood theoretically, it's a good foundation for our model. We have a degeneracy with Fe because we don't have a nucleosynthesis fact.}

In Section~\ref{sec:data} we describe the APOGEE data samples fit in this paper. In Section~\ref{sec:model} we outline our $K$-process model, the assumptions we make, and our implementation of the optimization routine. We apply the data-driven $K$-process model to the W22 APOGEE sample and present our findings in Section~\ref{sec:results_W22}. \ejg{Add ending discussion sections...}

\section{Data Samples}\label{sec:data}

In this paper, we employ stellar abundances from APOGEE DR17 \citep{abdurrouf2022}, part of SDSS-IV \citep{majewski2017}. The APOGEE survey obtains high-resolution ($R\sim22,500$) near-infrared (IR) observations \citep{wilson2019} for stars in the Galactic disk, halo, bulge, and nearby satellites/streams. Observations are taken with two nearly identical spectrographs on the 2.5m Sloan Foundation telescope \citep{wilson2019} at Apache Point Observatory in New Mexico and the 2.5m du Pont Telescope \citep{bowen1973} at the Las Companas Observatory in Chile. Spectral data are reduced and calibrated with the APOGEE data processing pipeline \citep{nidever2015}, after which stellar parameters and parameters are calculated with ASPCAP \citep[APOGEE Stellar Parameter and Chemical Abundance Pipeline][]{holtzman2015, garcia2016}. See \citet[][DR16]{jonsson2020} and Holtzman et al. (in prep., DR17) for a more detailed description of APOGEE data reduction and analysis, and \citep{zasowski2013, zasowski2017, santana2021} for a discussion of survey targeting.
\ejg{Is the DR17 paper out yet?}

APOGEE DR17 reports stellar parameters including $\teff$ and $\logg$ as well as 20 elemental abundances: C, \ion{C}{1}, N, O, Na, Mg, Al, Si, S, K, Ca, Ti, \ion{Ti}{2}, V, Cr, Mn, Fe, Co, Ni, and Ce for 657,135 stars. Among these elements and ions, some are measured more precisely than others. We exclude Ti from our analysis as there are large differences between the abundances derived from the \ion{Ti}{1} and \ion{Ti}{2} lines \citep{jonsson2020}. We also exclude P, as the P abundances are measured from a few very weak spectra features, and the abundance distribution displays strong artifacts \citep{jonsson2020}. Among the remaining elements, we note the following concerns:
\begin{itemize}
\itemsep0em
    \item Na - Spectral features are weak, one of the least precise elements in APOGEE (but matches optical data pretty well)
    \item S - Large abundance scatter in observations and no verification with optical data
    \item Cr - Significant systematic artifacts at super solar metallicity \citep{griffith2021a}
    \item Mn - Strong NLTE effects exist that are not taken into account
    \item Co - Abundance derived from one line, causing larger abundance scatter
    \item Ce - Abundance derived from one line. DR17 is the first time Ce abundances have been included in APOGEE data products. 
    \item Al,  V, and Mn require large offsets to place solar metallicity stars at solar [X/Fe].
\end{itemize}
\ejg{Will want to condense list for paper. Add a short description of some abundance systematics that we see re \citet{jonsson2020} and \citet{griffith2021a}}.

In this paper, we will explore two subsamples of APOGEE DR17 data: (1) the W22 sample, optimized to minimize statistical and systematic errors while probing a large portion of the Galactic disk, and (2) a sample with less stringent cuts, but better coverage of the low-metallicity ([Fe/H] $< -0.5$) disk.

\subsection{The W22 Sample}

W22 select a subset of APOGEE DR17 stars with the goal of minimizing statistical errors from poor observations and systematic errors from abundance trends with $\teff$ and/or $\logg$ while preserving a sufficient number of stars to conduct a meaningful statistical analysis across the Galactic disk. To remove poor quality data points, W22 require ASPCAP flags \texttt{STAR\_BAD} \texttt{NO\_ASPCAP\_RESULT} equal zero and remove all stars with \texttt{FE\_H\_FLAG} or \texttt{MG\_FE\_FLAG} set. They only include stars from the main survey sample (\texttt{EXTRATARG} = 0), and use named abundances (\texttt{X\_FE}), as recommended by \citet{jonsson2020}. In addition to these quality cute, W22 apply the following sample selection cuts:
\begin{itemize}
\itemsep0em
    \item $R=3-13$ kpc, $|Z| \leq 2$ kpc (distances from \citet{leung2019})
    \item $-0.75 \leq \mgh \leq 0.45$
    \item S/N $\geq 200$ for $\mgh > -0.5$ and S/N $\geq 100$ for $\mgh < -0.5$
    \item $\logg = 1-2.5$ dex
    \item $\teff = 4000-4600$ K
\end{itemize}
By selecting stars from a small range of $\logg$ and $\teff$, they eliminate the red clump stars and minimize stellar parameter-dependent abundance trends \citep[see][]{griffith2021a}. The cuts result in a sample of 34,410 stars.

W22 present abundance for Mg, O, Si, S, Ca, C+N, Na, Al, K, Cr, Fe, Ni, V, Mn, Co, and Ce. While the surface abundances of C and N differ from the stellar birth abundances for RGB stars due to the CNO processes and dredge up events \citep{iben1965, shetrone2019}, the total C+N abundance remains constant. W22 consider C+N as an element in their analysis, taking [(C+N)/H] to be 
\begin{equation}
    [\text{C+N}/\text{H}] = \log_{10}(10^{\text{[C/H]}+8.39} + 10^{\text{[N/H]}+7.78}) - \log_{10}(10^{8.39} + 10^{7.78})
\end{equation}
using logarithmic solar abundances for C (8.39) and N (7.78) from \citet{grevesse2007}.

\ejg{W22 use X/Fe error from APOGEE and don't convert it to X/Mg....Use C/Fe error for the C+N value because the ratio is mostly C... Need error discussion here and in next section.}

In the analysis of each element, X, stars with \texttt{X\_FE\_FLAG} are dropped (maximum of $\sim 560$ for Ce). We plot the distribution of abundances in [X/Mg] vs. [Mg/H] in Figure~\ref{fig:w22_xmg}. All elemental abundances have zero-point offsets applied and have been de-trended for correlations with $\teff$, as in W22 (see their Table 1).

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Figures/W22_xmg.pdf}
    \caption{W22 stellar abundance distributions. The number in the top right corner indicates the number of stars used in the analysis of that element. Elements are ordered by atomic number.}
    \label{fig:w22_xmg}
\end{figure}

\subsection{Our Sample}

\ejg{Not sure if this should go here or after we discuss the success of the K-process model on W22 data...}

While the W22 sample presented above is a good first test case of \name, we are interested in the behavior of our model when applied to a wider sample of data, in particular extending the metalicity range to lower values of [Mg/H]. To construct this sample, we apply the same quality cuts as taken by W22 (\texttt{STAR\_BAD} = 0, \texttt{NO\_ASPCAP\_RESULT} = 0, \texttt{FE\_H\_FLAG} = 0, \texttt{MG\_FE\_FLAG} = 0, and (\texttt{EXTRATARG} = 0). We add a cut for stars with the \texttt{ROTATION\_WARNING} flag set. We take the following sample selection cuts: 
\itemsep0em
\begin{itemize}
    \item S/N $\geq 100$
    \item $0 < \logg < 2.5$ dex
    \item $\teff = 4000-4600$ K
\end{itemize}
We make no cuts on the Galactic location, including stars from across the Galactic disk and halo, and make no cuts on the [Mg/H] values of our sample. The wider $\logg$ ranges expand the sample of giants. We select stars in the same small $\teff$ range as W22 to exclude the red clump \citep{vincenzo2021} and cool stars that show abundance artifacts \citep{jonsson2020}.
Our final sample includes 51,386 stars and extends to [Mg/H]=-2.2. As in W22, we only analyze [X/Mg] abundances for stars where the [X/Fe] value is unflagged. This has a minimal effect for most elements but reduces the sample size to 49,851 for Ce.

\ejg{Plus a magnitude cut as a proxy for distance? Right now I have H < 12, but I don't think that the cut impacts the global trends much...}

>>>>

% Abundance and stellar macros
\newcommand{\mgfe}[0]{[{\rm Mg/Fe}]}
\newcommand{\afe}[0]{[\alpha/{\rm Fe}]}
\newcommand{\femg}{[{\rm Fe}/{\rm Mg}]} 
\newcommand{\xmg}{[{\rm X}/{\rm Mg}]} 
\newcommand{\mgh}{[{\rm Mg}/{\rm H}]}
\newcommand{\feh}[0]{[{\rm Fe/H}]}
\newcommand{\xfe}{[{\rm X}/{\rm Fe}]}
\newcommand{\xh}{[{\rm X}/{\rm H}]}

\newcommand{\logg}{\log(g)}
\newcommand{\teff}{T_{\rm eff}}
\newcommand{\kpc}{\rm \; kpc}
\newcommand{\kel}{\rm \; K}
\newcommand{\msun}{M_{\odot}}
\newcommand{\rgc}{R_{\rm GC}}
\newcommand{\msunvice}{M_{\odot}/M_{\odot \rm formed}}
\newcommand{\MZAMS}{M_{\rm ZAMS}}

\newcommand{\qcc}{q_{{\rm CC,}j}^{Z}}
\newcommand{\qccFe}{q_{{\rm CC, Fe}}^{Z}}
\newcommand{\dqccFe}{dq_{{\rm CC, Fe}}^{Z}/dZ}
\newcommand{\qIa}{q_{{\rm Ia,}j}^{Z}}
\newcommand{\qIaFe}{q_{{\rm Ia,Fe}}^{Z}}
\newcommand{\Acc}{A^{\rm CC}_{i}}
\newcommand{\AIa}{A^{\rm Ia}_{i}}
\newcommand{\fcc}{f_{{\rm CC}, ij}}

\newcommand{\ejg}[1]{\textcolor{red}{EJG says: #1}}
\newcommand{\hogg}[1]{\textcolor{blue}{Hogg says: #1}}

@@

% text macros
\newcommand{\name}{\textsl{KPM}} 
\newcommand{\documentname}{\textsl{Article}}

% typesetting stuff
\shorttitle{a data-driven model for nucleosynthesis}
\shortauthors{Griffith \& Hogg}
\addtolength{\topmargin}{-0.4in} % trust in Hogg
\addtolength{\textheight}{0.7in} % trust in Hogg
\setlength{\parindent}{1.6em}
\renewcommand{\twocolumngrid}{} % once again
\renewcommand{\paragraph}[1]{\bigskip\par\noindent{\textbf{#1}}~---}
\sloppy\sloppypar\raggedbottom\frenchspacing % Trust Hogg

%\linespread{1.8}

@@

\title{\name: A flexible and data-driven $K$-process model for nucleosynthesis}

@@

\begin{abstract}\noindent % trust Hogg
Stellar surface abundances look like they are produced by two dominant processes, one prompt and one delayed.
We analyze the abundances of 14 elements in [X/Mg] vs. [Mg/H] space for 48,659 stars from APOGEE DR17 with a flexible, data-driven $K$-process model---dubbed \name.
In the fiducial model, where $K=2$, each elemental abundance of each star is described as the vector sum of a prompt (CCSN-like) and delayed (SNIa-like) process.
While prior work derives process vector components from median observed high-Ia and low-Ia abundance trends, we relax these assumptions and simultaneously fit all model parameters to the data, with only minimal constraints to keep the processes interpretable.
We find that our flexible method is able to recover the observed abundance bimodality and that this bimodality is visible to higher [Mg/H] in process amplitude space than in abundance space.
We compare fit parameters and statistics from this new method to those from prior work, finding that the two methods produce similar results, but that \name{} better predicts stellar abundances, especially for C+N and Mn.
We test the validity of the standard two-process model assumptions, such as that Fe is generated by a roughly half-half combination of prompt and delayed sources.
We find that the assumptions of prior works are generally confirmed, but that the inferred fractional contributions from each process are strongly dependent upon Fe assumptions.
We also explore third and fourth processes, calibrated to Ce and Mn, respectively. The additional processes improve the model's ability to predict the stellar abundances, demonstrating the model's ability to extend beyond $K=2$. \name{} enables future research to study the disk's two-dimensionallity, investigate age-abundance label correlations, fit non-bimodal abundance distributions with a multi-component nucleosynthetic model, and much more. \ejg{Add something about code availability.} 
\end{abstract}

@@

@@

After hydrogen, helium, lithium, and beryllium, all other naturally occurring elements are made in the stars, and the collisions of stars.
Stellar surface abundances---the abundances measured by taking a spectrum of a stellar photosphere---are thought to deliver a relatively unprocessed record of the element abundances in the gas from which the star formed \citep[though see e.g.,][]{pinsonneault2001, oh2018, souto2019, vincenzo2021b}.
These birth abundances were set by a combination of nucleosynthetic processes involved in making heavy atomic nuclei, and astrophysical processes involved in delivering atoms from stellar interiors to star-formation sites \citep[e.g.,][]{johnsonja2020}.
Thus nuclear physics and a wide swath of astrophysics are critically intertwined in our understanding of stellar surface abundances, motivating theoretical, experimental, and observational work.

At the present day, stellar surface abundances are not very well explained by \textsl{ab initio}, physics-driven models.
Yields vary from set to set, as they are dependent on progenitor properties and explosion assumptions \citep[e.g.,][]{rybizki2017, griffith2021b}. 
The wide parameter space of progenitor and supernovae models coupled with uncertainties in reaction rates and explosion physics hinder the creation of an accurate nucleosynthetic model from theory alone.
In the long run, it is incumbent upon us to understand these issues and correct the assumptions or calculations underlying our nucleosynthetic and astrophysical models.
In the short run, however, we gather data---tens of millions of abundance measurements on millions of stars in different astronomical surveys such as RAVE, SEGUE, LAMOST, Gaia-ESO, APOGEE/MWM, GALAH, and H3 \citep{steinmetz2006, yanny2009, gilmore2012, desilva2015, luo2015, majewski2017, conroy2019}.
This raises the question: \emph{Can we take a data-driven approach to nucleosynthesis?}

In this \documentname{}, we build a purely data-driven model for the surface element abundances observed in stars.
We treat each star as being a linear combination of nucleosynthetic processes, beginning with one that is primarily responsible for the $\alpha$-element Mg \citep[prompt enrichment, such as core collapse supernovae or CCSN, e.g.,][]{andrews2017}, and one that is primarily \emph{not} responsible for Mg (delayed enrichment, such as Type-Ia supernovae or SNIa).
Beyond these up-front assumption, we try to be agnostic about how the elements are produced.

We build upon the work of \citet[][hereafter G22]{griffith2019, griffith2022} and \citet[][hereafter W19, W22]{weinberg2019, weinberg2022}, who used the bimodality in [Mg/Fe] vs. [Fe/H]\footnote{Where $[{\rm X}/{\rm Y}] = \log({\rm X}/{\rm Y}) - \log({\rm X}/{\rm Y})_{\odot}$ and $({\rm X}/{\rm Y})_{\odot}$ is the solar abundance ratio.} \citep[e.g.,][]{fuhrmann1998, bensby2003, adibekyan2012} to separate stars into populations with high and low SNIa enrichment. Using the median [X/Mg] vs. [Mg/H] abundance trends, these works explain data from the \textsl{GALAH}\footnote{GALAH = GALactic Archaeology with HERMES.} and SDSS-IV \textsl{APOGEE}\footnote{APOGEE = Apache Point Observatory Galactic Evolution Experiment, part of the Sloan Digital Sky Survey} surveys, respectively, with a two-process model. Because the median abundance trends in [X/Mg] vs. [Mg/H] space are largely insensitive to aspects of chemical evolution, such as outflows and variations in star formation history \citep{weinberg2019}, the population abundance trends are set by the nucleosynthetic processes and can be used to empirically constrain Galactic enrichment.

These works, as well as \citet{ting2022}, find that the Milky Way stellar abundances are well fit two components, down to residuals of 0.01 to 0.02 dex. Simultaneously, \citet{frankel2018} and \citet{ness2022} have found that disk abundances are also well described by a two component model of birth radius and age. Correlations between two-process model parameters and stellar ages and kinematics (W22) suggest that these two 2-dimensional models are somehow interconnected. \ejg{Add bit on why two processes dominate---two main channels v.s. mixing in the disk.}

Beyond standard CCSN and SNIa enrichment many elements have contributions from additional nucleosynthetic processes, such as the rapid ($r$) and slow ($s$) neutron capture processes \citep[e.g.,][]{arlandini1999, bisterzo2014} in asymptotic giant branch (AGB) stars \citep[e.g.,][]{simmerer2004, karakas2016} and merging neutron stars \citep[e.g.,][]{kilpatrick2017}, or atypical supernovae explosion \citep[e.g.,][]{nomoto2013}. After predicting stellar abundances from $\feh$ and $\mgfe$, \citet{ting2022} identify correlated abundance residuals that are unexplained by observational uncertainties, indicative of additional nucleosynthetic processes that standard disk CCSN and SNIa enrichment cannot explain. Results from G22 and W22 support this conclusion, and both works attempt to add additional processes to their models to account for non-CCSN and non-SNIa enrichment, though in a very restrictive manner. Other sources of abundance scatter, such as stochastic sampling of the IMF, IMF variations, and bursty star formation history could also cause deviations away from a two-process model \citep{belokurov2018, griffith2023}.

To date, survey abundances have not been fully exploited to create a data-driven model of nucleosynthesis. While works such as \citet{ting2012}, \citet{casey2019}, and \citet{ratcliffe2022} effectively use clustering algorithms to identify elements with like sources and reduce abundance dimensionality, the results are difficult to translate into a model of nucleosynthesis. Clustering components can be linked to nucleosynthesis sources and enrichment history, but have not yet been be used to describe the enrichment of a single star.

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data. In the $K$ Process Model (\name{}), we find the intersection between reliable facts about nucleosynthesis and good abundance measurements to build an edifice of Galactic enrichment. The parameters output by our model can be used as de-noised abundance label to sharpen relationships between abundances and stellar parameters.  \ejg{Add more about high-quality metallicity labels? Hogg to edit this paragraph.}

This paper is organized as follows. In Section~\ref{sec:model} we present the assumptions and the implementation of \name{}. In Section~\ref{sec:data} we describe the APOGEE data sample employed in this paper.  We apply \name{} to the APOGEE data in Section~\ref{sec:fiducial} and compare our results to those of W22 in Section~\ref{subsec:w22}. In Section~\ref{sec:variations} we explore variations from the fiducial model, changing our assumptions about Fe production as well as the number of model components. Finally, we discuss and summarize our results in Section~\ref{sec:discussion}.

\section{The $K$-Process Model}\label{sec:model}

As in W22 and G22, we propose that all stellar abundances can be generated by a combination of $K$ nucleosynthetic processes.
In this picture, each element has $K$ metallicity-dependent process vector components that are shared across the full stellar sample, while each star individually has $K$ process amplitudes, which apply across all elements, such that
\begin{equation}\label{eq:mij_k}
    m_{ij} = \log_{10} \, \sum^K_{k=1} A_i^k \, q_{k,j}^Z ~,
\end{equation}
where stars are labeled by $i$ and elements are labeled by $j$.
Each star $i$ has $K$ process amplitudes ($A^k_i$)
and each element $j$ has $K$ process vector components ($q_{k,j}^{Z}$).
The $m_{ij}$ value is the model expected logarithmic abundance of element $j$ in star $i$. In this model, we put forth the following set of assumptions:

\paragraph{1. $K$ processes}
All elements on the periodic table are produced by a combination of nucleosynthetic processes such as CCSN, SNIa, AGB stars, and merging neutron stars \citep{johnsonja2020}. The majority of $\alpha$, light odd-Z, and Fe-peak elements (the elements observed by APOGEE) are dominantly produced by $K=2$ sources, with one being a prompt process or mix of prompt processes, and one being a delayed process or a mix of delayed processes. This is substantiated by theoretical yields \citep[e.g.,][]{anderson2019, rybizki2017} and past successful data-driven models (e.g., G22, W22, \citealp{ting2022}). In this paper we therefore assume that $K \geq 2$, though \name{} could be implemented with $K=1$. \hogg{This assumption says too much; we could dial this back.}

\paragraph{2. Linearity}
At every metallicity, the (X/H) abundances of a star can be expressed as a linear combination of $K$ processes.
These processes themselves will depend on metallicity, but a linear sum is sufficient to explain all element abundances at any overall metallicity.
Because yields depend on stellar masses and progenitor structure, and because different stars can get to their metallicities by different histories, this assumption must be at least slightly wrong in detail.

\paragraph{3. Non-negativity}
All process vector components for all elements are non-negative and all process amplitudes are non-negative.
This assumption implies that the elements considered here are only produced, and not ever destroyed, by the $K$ processes (relative to hydrogen).
This makes the model similar to a non-negative matrix factorization (\citealt{nnmf} \ejg{Hogg add citation}). In \name{}, this assumption is enforced by requiring that the process vector components and amplitudes are always greater than or equal to zero, such that
\begin{equation} \label{eq:constraint_zero}
    q_{k,j}^Z \geq 0 \; \forall \; Z,k,j \quad \text{and} \quad A_i^k \geq 0 \; \forall \; k,i ~.
\end{equation}

\paragraph{4. Mg production}
All Mg is produced in a prompt process and no other processes contribute to its production.
This is substantiated by theoretical yields where Mg is purely produced by prompt CCSN \citep[e.g.,][]{anderson2019, rybizki2017}.
This assumption (along with non-negativity) breaks a set of symmetries in the process space and makes the processes quasi-interpretable in terms of nucleosynthesis sources.
Because such a prompt process is likely dominated by CCSN \citep[e.g.,][]{andrews2017}, we label the first process with ``CC''.

In \name{}, this assumption is enforced by fixing the Mg process vector components such that
\begin{equation}\label{eq:qcc_mg_solar}
    q_{{\rm CC, Mg}}^{\,Z_{\odot}} = 1, \quad q_{k>1,{\rm Mg}}^{\,Z_{\odot}} = 0
\end{equation}
at solar metallicity and 
\begin{equation}\label{eq:qcc_mg_z}
    q_{{\rm CC, Mg}}^{\,Z} = q_{{\rm CC, Mg}}^{\,Z_{\odot}}, \quad q_{k>1, {\rm Mg}}^{\,Z} = q_{k>1, {\rm Mg}}^{\,Z_{\odot}}
\end{equation}
at all other metallicities.

\paragraph{5. Fe production}
Fe is produced through a combination of a prompt and delayed process. Because the delayed process is likely dominated by SNIa \citep[e.g.,][]{andrews2017}, we label the delayed process with ``Ia''\footnote{While other enrichment channels with similar timescales may be included in the respective processes, the ``CC'' and ``Ia'' naming convention conforms to the choices in W22 and G22, and avoids the possible confusion of process numbers (1 and 2) with supernovae type (II and Ia).}.
While the prompt process constraint (Mg) is grounded in nucleosynthesis theory, there is no equivalent nucleosynthesis fact to constrain the delayed process.
To break model degeneracies, we also fix the Fe process vector components such that

\begin{equation}\label{eq:qcc_fe_solar}
    q_{{\rm CC, Fe}}^{\,Z_{\odot}} = 0.4, \quad q_{{\rm Ia, Fe}}^{\,Z_{\odot}} = 1 - q_{{\rm CC, Fe}}^{\,Z_{\odot}} \quad q_{k>2,{\rm Fe}}^{\,Z_{\odot}} = 0
\end{equation}
at solar metallicity and 
\begin{equation}\label{eq:qcc_fe_z}
    q_{{\rm CC, Fe}}^{\,Z} = q_{{\rm CC, Fe}}^{\,Z_{\odot}},  \quad 
    q_{{\rm Ia, Fe}}^{\,Z} = q_{{\rm Ia, Fe}}^{\,Z_{\odot}} \quad 
    q_{k>2, {\rm Fe}}^{\,Z} = q_{k>2, {\rm Fe}}^{\,Z_{\odot}}
\end{equation}
at all other metallicities.
This assumption places a star with purely prompt enrichment on the low-metallicity $\mgfe$ plateau near 0.4 dex, in agreement with APOGEE observations but in contention with recent results from \citet{conroy2022} which places the $\mgfe$ plateau near 0.6 dex.
We explore the impact of different $\qccFe$ assumptions in Section~\ref{subsec:qccFe}.

\paragraph{6. Metallicity dependence}
We permit the process vector components for all elements other than Mg and Fe to float as a function of metallicity.
The variation is parameterized by a linear spline in log-process space, attached to a set of variable control points. We assume that a particular set of hard-coded knots are sufficient to capture the metallicity dependence. \hogg{This could be mentioned in passing in the discussion where you talk about model improvements.}

\paragraph{7. APOGEE abundances and uncertainties}
We assume that the APOGEE abundances and uncertainties can be used for this project.
This is not the same as assuming that they are correct, but rather that it is possible and useful to build an interpretable model to explain them. We describe the potential data systematics in Section~\ref{sec:data}.
The APOGEE errors are likely slightly underestimated for some element measurements, so we add softening parameter $Q$ to inflate the errors (see Equation~\ref{eq:inflate_ivar}). \hogg{Make this more positive about apogee}
